---
title: "Hands-on Exercise 1"
author: "Sathvika Subramanian"
date: "January 16, 2025"
date-modified: "last-modified"
---

# 1. Getting Started

## 1.1 Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will launched into R.

```{r}
pacman::p_load(tidyverse)
```

## 1.2 Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

1.  Year end examination grades of a cohort of primary 3 students from a local school.

2.  There are a total of seven attributes. Four of them are categorical data type and the other three are in continuous data type.

3.  The categorical attributes are: ID, CLASS, GENDER and RACE. The continuous attributes are: MATHS, ENGLISH and SCIENCE.

# 2. R graphics vs ggplot

```{r}
hist(exam_data$MATHS)
```

```{r}
#| code-fold: show

ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="light blue") +
  ggtitle("Distribution of Maths scores")
```

# 3. Essential Grammatical Elements in ggplot2

## 3.1 Data

```{r}
ggplot(data=exam_data)
```

A blank canvas appears. ggplot() initializes a ggplot object.

## 3.2 Aesthetic mappings

ggplot includes the x-axis and the axis’s label.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

## 3.3 Geom

### 3.3.1 Geometric Objects: geom_bar

Plots a bar chart of RACE by using geom_bar()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

Plots a bar chart of GENDER by using geom_bar()

```{r}
ggplot(data=exam_data, 
       aes(x=GENDER)) +
  geom_bar()
```

### 3.3.2 Geometric Objects: geom_dotplot

Dot plot of MATHS from exam_data

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

For the above graph y is very misleading, thus scale_y_continuous() is used to turn off the y-axis, and binwidth argument is used to change the binwidth to 2.5.

```{r}
#| code-fold: show

ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

### 3.3.3 Geometric Objects: geom_histogram()

Creates a simple histogram by using values in MATHS field of exam_data.

```{r setup, warning=FALSE, message=FALSE}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       
```

Creates a simple histogram by using values in SCIENCE field of exam_data.

```{r warning=FALSE, message=FALSE}
ggplot(data=exam_data, 
       aes(x = SCIENCE)) +
  geom_histogram()       
```

### 3.3.4 Modifying a geometric object by changing geom()

```{r}
#| code-fold: show

ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light green")  
```

### 3.3.5 Modifying a geometric object by changing aes()

Changes the interior colour of the histogram (i.e. fill) by using sub-group of aesthetic()

```{r}
#| code-fold: show

 ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="black")
```

### 3.3.6 Geometric Objects: geom-density()

Plots the distribution of Maths scores in a kernel density estimate plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

Plots two kernel density lines by using colour or fill arguments of aes()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

### 3.3.7 Geometric Objects: geom_boxplot

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

Notch plot

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

### 3.3.8 Geometric Objects: geom_violin

Plots the distribution of Maths score by gender in violin plot

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

### 3.3.9 Geometric Objects: geom_point()

Plots a scatterplot showing the Maths and English grades of pupils by using geom_point()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

Plots the data points on the boxplots by using both geom_boxplot() and geom_point().

```{r}
#| code-fold: show

ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

## 3.4 Stat

### 3.4.1 Working with stat()

The boxplots below are incomplete because the positions of the means were not shown.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

### 3.4.2 Working with stat - the stat_summary() method

Adds mean values by using stat_summary() function and overriding the default geom.

```{r}
#| code-fold: show

ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="orange",        
               size=4)               
```

### 3.4.3 Working with stat - the geom() method

Adds mean values by using geom\_() function and overriding the default stat.

```{r}
#| code-fold: show

ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="orange",          
             size=4)          
```

### 3.4.4 Adding a best fit curve on a scatterplot?

geom_smooth() is used to plot a best fit curve on the scatterplot.

```{r warning=FALSE, message=FALSE}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

The default smoothing method is overridden

```{r warning=FALSE, message=FALSE}
#| code-fold: show

ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

## 3.5 Facets

### 3.5.1 Working with facet_wrap()

The code chunk below plots a trellis plot using facet-wrap()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

### 3.5.2 facet_grid() function

Plots a trellis plot using facet_grid().

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## 3.6 Coordinates

### 3.6.1 Working with Coordinate

By the default, the bar chart of ggplot2 is in vertical form.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

Flips it into vertical bar chart

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

### 3.6.2 Changing the y- and x-axis range

```{r warning=FALSE, message=FALSE}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

Setting fixed range for x and y-axis as the ranges are misleading in the above chart

```{r warning=FALSE, message=FALSE}
#| code-fold: show

ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## 3.7 Themes

### 3.7.1 Working with theme

Theme gray

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

Theme classic

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

Theme minimal

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

Theme light

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_light()
```

Theme bw

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_bw()
```
