---
title: "Hands on Exercise 5-4"
author: "Sathvika Subramanian"
date: "February 8, 2025"
date-modified: "last-modified"
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
---

# 1. Visual Multivariate Analysis with Parallel Coordinates Plot


# 2. Visual Multivariate Analysis with Parallel Coordinates Plot

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```


# 3. Data Preparation

```{r}
wh <- read_csv("data/WHData-2018.csv")
```


# 4. Plotting Static Parallel Coordinates Plot

## 4.1 Plotting a simple parallel coordinates

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```

## 4.2 Plotting a parallel coordinates with boxplot

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

## 4.3 Parallel coordinates with facet

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

## 4.4 Rotating x-axis text label

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

## 4.5 Adjusting the rotated x-axis text label

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```


# 5.Plotting Interactive Parallel Coordinates Plot: parallelPlot methods

## 5.1 The basic plot

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```

## 5.2 Rotate axis label

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

## 5.3 Changing the colour scheme

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

## 5.4 Parallel coordinates plot with histogram

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```
