---
title: "Hands on Exercise 5-5"
author: "Sathvika Subramanian"
date: "February 8, 2025"
date-modified: "last-modified"
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
---

# 1. Treemap Visualisation with R

# 2. Installing and Launching R Packages

```{r}
pacman::p_load(treemap, treemapify, tidyverse) 
```

# 3. Data Wrangling

## 3.1 Importing the data set

```{r}
realis2018 <- read_csv("data/realis2018.csv")
```

## 3.2 Grouped summaries without the Pipe

```{r}
realis2018_grouped <- group_by(realis2018, `Project Name`,
                               `Planning Region`, `Planning Area`, 
                               `Property Type`, `Type of Sale`)
realis2018_summarised <- summarise(realis2018_grouped, 
                          `Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
                          `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
                          `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE), 
                          `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```


## 3.3 Grouped summaries with the pipe

```{r}
realis2018_summarised <- realis2018 %>% 
  group_by(`Project Name`,`Planning Region`, 
           `Planning Area`, `Property Type`, 
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE), 
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

## 3.4 Designing Treemap with treemap Package

### 3.4.1 Designing a static treemap

```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
```

### 3.4.2 Using the basic arguments

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 3.4.3 Working with vColor and type arguments

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type = "value",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 3.4.4 Colours in treemap package - The “value” type treemap

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 3.4.5 The “manual” type treemap

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 3.4.6 Treemap Layout - Working with algorithm argument

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 3.4.7 Using sortID

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "pivotSize",
        sortID = "Median Transacted Price",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```


# 4. Designing Treemap using treemapify Package

## 4.1 Designing a basic treemap

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start = "bottomleft") + 
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

## 4.2 Defining hierarchy

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") + 
  geom_treemap()
```

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap()
```

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```

# 5. Designing Interactive Treemap using d3treeR

## 5.1 Installing d3treeR package

```{r}
library(devtools)
install_github("timelyportfolio/d3treeR")
```

```{r}
library(d3treeR)
```

## 5.2 Designing An Interactive Treemap

```{r}
tm <- treemap(realis2018_summarised,
        index=c("Planning Region", "Planning Area"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        title="Private Residential Property Sold, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

```{r}
d3tree(tm,rootname = "Singapore" )
```
