---
title: "In class Exercise 7"
author: "Sathvika Subramanian"
date: "March 1, 2025"
date-modified: "last-modified"
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
---

# 1. Getting Started

```{r}
pacman::p_load(tidyverse, tsibble, feasts, fable, seasonal)
```

## 1.1 Importing Data

```{r}
ts_data <- read_csv(
  "data/visitor_arrivals_by_air.csv")
```

converting from character to date field
```{r}
ts_data$`Month-Year` <- dmy(
  ts_data$`Month-Year`)
```

## 1.2 Conventional base ts object versus tibble object

```{r}
ts_data
```

```{r}
ts_data_ts <- ts(ts_data)       
head(ts_data_ts)
```

## 1.3 Converting tibble object to tsibble object

```{r}
ts_tsibble <- ts_data %>%
  mutate(Month = yearmonth(`Month-Year`)) %>%
  as_tsibble(index = `Month`)
```

```{r}
ts_tsibble
```


# 2.Visualising Time-series Data

```{r}
ts_longer <- ts_data %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```

## 2.1 Visualising single time-series: ggplot2 methods

```{r}
ts_longer %>%
  filter(Country == "Malaysia") %>%
  ggplot(aes(x = `Month-Year`, 
             y = Arrivals))+
  geom_line(size = 0.5)
```

## 2.2 Plotting multiple time-series data with ggplot2 methods

```{r}
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals,
           color = Country))+
  geom_line(size = 0.5) +
  theme(legend.position = "bottom", 
        legend.box.spacing = unit(0.5, "cm"))
```

```{r}
#| echo: false
#| fig-height: 12

ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals))+
  geom_line(size = 0.5) +
  facet_wrap(~ Country,
             ncol = 3,
             scales = "free_y") +
  theme_bw()
```


# 3. Visual Analysis of Time-series Data

```{r}
tsibble_longer <- ts_tsibble %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```


## 3.1 Visual Analysis of Seasonality with Seasonal Plot

```{r}
tsibble_longer %>%
  filter(Country == "Italy" |
         Country == "Vietnam" |
         Country == "United Kingdom" |
         Country == "Germany") %>% 
  gg_season(Arrivals)
```

## 3.2 Visual Analysis of Seasonality with Cycle Plot

```{r}
tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  autoplot(Arrivals) + 
  facet_grid(Country ~ ., scales = "free_y")
```

```{r}
#| fig-height: 6

tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  gg_subseries(Arrivals)
```
 
# 4. Time series decomposition

## 4.1 Single time series decomposition

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  ACF(Arrivals) %>% 
  autoplot()
```

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  PACF(Arrivals) %>% 
  autoplot()
```


## 4.2 Multiple time-series decomposition

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam" |
         `Country` == "Italy" |
         `Country` == "United Kingdom" |
         `Country` == "China") %>%
  ACF(Arrivals) %>%
  autoplot()
```

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam" |
         `Country` == "Italy" |
         `Country` == "United Kingdom" |
         `Country` == "China") %>%
  PACF(Arrivals) %>%
  autoplot()
```


## 4.3 Composite plot of time series decomposition

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  gg_tsdisplay(Arrivals)
```


# 5.Visual STL Diagnostics

## 5.1 Visual STL diagnostics with feasts

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  model(stl = STL(Arrivals)) %>%
  components() %>%
  autoplot()
```

## 5.2 Classical Decomposition with feasts

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  model(
    classical_decomposition(
      Arrivals, type = "additive")) %>%
  components() %>%
  autoplot()
```


# 6. Visual Forecasting

## 6.1 Time Series Data Sampling

```{r}
vietnam_ts <- tsibble_longer %>%
  filter(Country == "Vietnam") %>% 
  mutate(Type = if_else(
    `Month-Year` >= "2019-01-01", 
    "Hold-out", "Training"))
```

```{r}
vietnam_train <- vietnam_ts %>%
  filter(`Month-Year` < "2019-01-01")
```


## 6.2 Exploratory Data Analysis (EDA): Time Series Data

```{r}
vietnam_train %>%
  model(stl = STL(Arrivals)) %>%
  components() %>%
  autoplot()
```

## 6.3 Fitting forecasting models

### 6.3.1 Fitting Exponential Smoothing State Space (ETS) Models: fable methods

```{r}
ETS(y ~ error(c("A", "M")) 
    + trend(c("N", "A", "Ad")) 
    + season(c("N", "A", "M")))
```

### 6.3.2 Fitting a simple exponential smoothing (SES)

```{r}
fit_ses <- vietnam_train %>%
  model(ETS(Arrivals ~ error("A") 
            + trend("N") 
            + season("N")))
fit_ses
```

### 6.3.3 Examine Model Assumptions

```{r}
gg_tsresiduals(fit_ses)
```

### 6.3.4 The model details

```{r}
fit_ses %>%
  report()
```

### 6.3.5 Trend methods

```{r}
vietnam_H <- vietnam_train %>%
  model(`Holt's method` = 
          ETS(Arrivals ~ error("A") +
                trend("A") + 
                season("N")))
vietnam_H %>% report()
```

### 6.3.6 Damped Trend methods 

```{r}
vietnam_HAd <- vietnam_train %>%
  model(`Holt's method` = 
          ETS(Arrivals ~ error("A") +
                trend("Ad") + 
                season("N")))
vietnam_HAd %>% report()
```

### 6.3.7 Checking for results

```{r}
gg_tsresiduals(vietnam_H)
```

```{r}
gg_tsresiduals(vietnam_HAd)
```

## 6.4 Fitting ETS Methods with Season: Holt-Winters

```{r}
Vietnam_WH <- vietnam_train %>%
  model(
    Additive = ETS(Arrivals ~ error("A") 
                   + trend("A") 
                   + season("A")),
    Multiplicative = ETS(Arrivals ~ error("M") 
                         + trend("A") 
                         + season("M"))
    )

Vietnam_WH %>% report()
```

## 6.5 Fitting multiple ETS Models

```{r}
fit_ETS <- vietnam_train %>%
  model(`SES` = ETS(Arrivals ~ error("A") + 
                      trend("N") + 
                      season("N")),
        `Holt`= ETS(Arrivals ~ error("A") +
                      trend("A") +
                      season("N")),
        `damped Holt` = 
          ETS(Arrivals ~ error("A") +
                trend("Ad") + 
                season("N")),
        `WH_A` = ETS(
          Arrivals ~ error("A") + 
            trend("A") + 
            season("A")),
        `WH_M` = ETS(Arrivals ~ error("M") 
                         + trend("A") 
                         + season("M"))
  )
```

## 6.6 The model coefficient

```{r}
fit_ETS %>%
  tidy()
```

## 6.7 Model Comparison

```{r}
fit_ETS %>% 
  report()
```

## 6.8 Forecasting future values

```{r}
fit_ETS %>%
  forecast(h = "12 months") %>%
  autoplot(vietnam_ts, 
           level = NULL)
```

## 6.9 Fitting ETS Automatically

```{r}
fit_autoETS <- vietnam_train %>%
  model(ETS(Arrivals))
fit_autoETS %>% report()
```

## 6.10 Fitting Fitting ETS Automatically

```{r}
gg_tsresiduals(fit_autoETS)
```

## 6.11  Forecast the future values

```{r}
fit_autoETS %>%
  forecast(h = "12 months") %>%
  autoplot(vietnam_train)
```

## 6.12 Visualising AutoETS model with ggplot2

```{r}
fc_autoETS <- fit_autoETS %>%
  forecast(h = "12 months")

vietnam_ts %>%
  ggplot(aes(x=`Month`, 
             y=Arrivals)) +
  autolayer(fc_autoETS, 
            alpha = 0.6) +
  geom_line(aes(
    color = Type), 
    alpha = 0.8) + 
  geom_line(aes(
    y = .mean, 
    colour = "Forecast"), 
    data = fc_autoETS) +
  geom_line(aes(
    y = .fitted, 
    colour = "Fitted"), 
    data = augment(fit_autoETS))
```


# 7. AutoRegressive Integrated Moving Average(ARIMA) Methods for Time Series Forecasting: fable (tidyverts) methods

## 7.1 Visualising Autocorrelations: feasts methods

```{r}
vietnam_train %>%
  gg_tsdisplay(plot_type='partial')
```

## 7.2 Visualising Autocorrelations: feasts methods

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  ACF(Arrivals) %>% 
  autoplot()
```

```{r}
tsibble_longer %>%
  filter(`Country` == "United Kingdom") %>%
  ACF(Arrivals) %>% 
  autoplot()
```


## 7.3 Differencing: fable methods

### 7.3.1 Trend differencing

```{r}
tsibble_longer %>%
  filter(Country == "Vietnam") %>%
  gg_tsdisplay(difference(
    Arrivals,
    lag = 1), 
    plot_type='partial')
```

### 7.3.2 Seasonal differencing

```{r}
tsibble_longer %>%
  filter(Country == "Vietnam") %>%
  gg_tsdisplay(difference(
    Arrivals,
    difference = 12), 
    plot_type='partial')
```

## 7.4 Fitting ARIMA models manually: fable methods

```{r}
fit_arima <- vietnam_train %>%
  model(
    arima200 = ARIMA(Arrivals ~ pdq(2,0,0)),
    sarima210 = ARIMA(Arrivals ~ pdq(2,0,0) + 
                        PDQ(2,1,0))
    )
report(fit_arima)
```

## 7.5 Fitting ARIMA models automatically: fable methods

```{r}
fit_autoARIMA <- vietnam_train %>%
  model(ARIMA(Arrivals))
report(fit_autoARIMA)
```

## 7.6 Model Comparison

```{r}
bind_rows(
    fit_autoARIMA %>% accuracy(),
    fit_autoETS %>% accuracy(),
    fit_autoARIMA %>% 
      forecast(h = 12) %>% 
      accuracy(vietnam_ts),
    fit_autoETS %>% 
      forecast(h = 12) %>% 
      accuracy(vietnam_ts)) %>%
  select(-ME, -MPE, -ACF1)
```

## 7.7 Forecast Multiple Time Series

```{r}
ASEAN <- tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Malaysia" |
         Country == "Indonesia" |
         Country == "Thailand" |
         Country == "Philippines")
```

```{r}
ASEAN_train <- ASEAN %>%
  mutate(Type = if_else(
    `Month-Year` >= "2019-01-01", 
    "Hold-out", "Training")) %>%
  filter(Type == "Training")
```

## 7.8 Fitting Mulltiple Time Series

```{r}
ASEAN_fit <- ASEAN_train %>%
  model(
    ets = ETS(Arrivals),
    arima = ARIMA(Arrivals)
  )
```

## 7.9 Examining Models

```{r}
ASEAN_fit %>%
  glance()
```

## 7.10 Extracintg fitted and residual values

```{r}
ASEAN_fit %>%
  augment()
```

## 7.11 Comparing Fit Models

```{r}
ASEAN_fit %>%
  accuracy() %>%
  arrange(Country)
```

## 7.12 Forecast Future Values

```{r}
ASEAN_fc <- ASEAN_fit %>%
  forecast(h = "12 months")
```

## 7.13 Visualising the forecasted values

```{r}
ASEAN_fc %>%
  autoplot(ASEAN)
```
