---
title: "Hands-on_Ex4-1"
author: "Sathvika Subramanian"
date: "February 5, 2025"
date-modified: "last-modified"
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
---

# Visualising Distribution

# 1. Getting Started

## 1.1 Installing and loading the packages

```{r}
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, tidyverse)
```

## 1.2 Data Import

```{r}
exam <- read_csv("data/Exam_data.csv")
```


# 2. Visualising Distribution with Ridgeline Plot

## 2.1 Plotting ridgeline graph: ggridges method

```{r}
#| code-fold: show

ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB",.3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0,0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

## 2.2 Varying fill colors along the x-axis

```{r}
#| code-fold: show

ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0,0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion (add = c(0.2, 2.6))) + 
  theme_ridges()
```

## 2.3 Mapping the probabilities directly onto colour

```{r}
#| code-fold: show

ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```

## 2.4 Ridgeline plots with quantile lines

```{r}
#| code-fold: show

ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(stat(quantile)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```

```{r}
#| code-fold: show

ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(stat(quantile)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = c(0.025, 0.975)
    )+
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")) +
  theme_ridges()
```


# 3. Visualising Distribution with Raincloud Plot

## 3.1 Plotting a Half Eye graph

```{r}
#| code-fold: show

ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```

## 3.2 Adding the boxplot with geom_boxplot()

```{r}
#| code-fold: show

ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)
```

## 3.3 Adding the Dot Plots with stat_dots()

```{r}
#| code-fold: show

ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left",
            justification = 1.2,
            binwidth = .5,
            dotsize = 2)

```

## 3.4 Finishing touch

```{r}
#| code-fold: show

ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left",
            justification = 1.2,
            binwidth = .5,
            dotsize = 1.5
            ) +
  coord_flip() +
  theme_classic()
```